---
title: "Temperature and mortality in Chicago, 1987-2000: a DLNM and case-crossover demonstration"
author: "Raymond Reuel Wayesu"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  github_document:
    toc: true
    toc_depth: 2
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
rd  <- function(n) readRDS(file.path("..", "outputs", paste0(n, ".rds")))
fig <- function(f) knitr::include_graphics(file.path("..", "outputs", "figures", f))
source(file.path("..", "R", "00_config.R"))
dat <- rd("dat"); tab1 <- rd("tab1"); ts <- rd("fit_timeseries"); cc <- rd("fit_casecrossover")
af  <- rd("attributable"); se <- rd("sensitivity"); it <- rd("interaction")
x <- dat[[spec$exposure]]
ci <- function(e, l, h, d = 2) sprintf(paste0("%.", d, "f (95%% CI %.", d, "f to %.", d, "f)"), e, l, h)
rr <- function(tab, p) { r <- tab[tab$percentile == p, ]; ci(r$rr, r$lo, r$hi) }
im <- function(m) { r <- it$table[it$table$measure == m, ]; ci(r$estimate, r$lo, r$hi) }
afr <- function(k) { r <- af$af[k, ]; sprintf("%.2f%% (95%% eCI %.2f to %.2f)", r$af, r$af_lo, r$af_hi) }
best <- se$sens[which.min(se$sens$qaic), ]
```

> **What this is.** A small, fully reproducible methods demonstration on an open
> teaching dataset. It reproduces the standard environmental-epidemiology toolkit
> for short-term temperature effects. It is **not** original research, makes no
> new scientific claim, and uses no restricted, clinical or HeatShield user data.

## 1. Question

How does daily mean temperature relate to all-cause mortality over the following
three weeks, where is the temperature of minimum mortality, how much of the death
toll is attributable to cold and to heat, and does the answer depend on the study
design or on defensible modelling choices?

## 2. Data

`chicagoNMMAPS`, distributed with the **dlnm** R package (GPL >= 2): `r format(nrow(dat), big.mark = ",")`
consecutive days, `r min(dat$date)` to `r max(dat$date)`, `r format(sum(dat$death), big.mark = ",")`
deaths. The data are loaded from the package at run time and are not stored in this repository.

```{r}
knitr::kable(tab1, caption = "Table 1. Daily series: missing values, mean and percentiles.")
```

## 3. Methods

**Exposure-lag-response.** A distributed lag non-linear model (Gasparrini et al. 2010;
Gasparrini 2011). The cross-basis combines a natural cubic spline of temperature (knots at the
`r paste0(100 * spec$var_knots_p, "th", collapse = ", ")` percentiles) with a natural cubic
spline over lags 0-`r spec$max_lag` days (`r spec$lag_nknots` knots equally spaced on the log scale).

**Design 1: time series.** Quasi-Poisson regression with a natural spline of time
(`r spec$df_per_year` df per year) and day of week.

**Design 2: time-stratified case-crossover.** Conditional quasi-Poisson regression with strata of
year x month x day of week (`r cc$n_strata` strata), which is equivalent to the conditional-logistic
case-crossover analysis but allows overdispersion (Armstrong et al. 2014). The cross-basis is
identical, so differences between designs reflect the design alone.

**Minimum mortality temperature (MMT).** Minimum of the overall cumulative curve, searched between
the `r 100 * spec$mmt_search_p[1]`st and `r 100 * spec$mmt_search_p[2]`th temperature percentiles; empirical CI
from `r format(spec$n_sim, big.mark = ",")` Monte Carlo draws of the coefficients.

**Attributable burden.** Backward-perspective attributable fraction and number
(Gasparrini & Leone 2014) relative to the MMT, with empirical CIs. The estimator was written
independently for this project and reproduces the authors' reference function `attrdl()` to
machine precision (`R/99_validate_attributable.R`).

**Sensitivity.** `r length(spec$sens_df)` x `r length(spec$sens_lag)` grid of df per year and maximum lag, all fitted
to identical rows and compared by QAIC.

**Interaction.** Warm season only (May-September). The heat DLNM (lag 0-`r spec$int_max_lag`) is allowed to
differ between low- and high-ozone days; interaction is reported on the multiplicative scale and
on the additive scale (RERI), with Monte Carlo CIs from the joint coefficient distribution.

All modelling choices live in `R/00_config.R`. The main specification was fixed **a priori**,
following common practice in multi-city studies (Gasparrini et al. 2015), not selected by QAIC.

## 4. Results

### 4.1 Overall association and the two designs

The MMT was **`r sprintf("%.1f", ts$mmt$mmt)` C** (95% eCI `r sprintf("%.1f to %.1f", ts$mmt$ci[1], ts$mmt$ci[2])`),
the `r round(100 * ts$mmt$pct)`th percentile of the temperature distribution. Overdispersion was modest
(dispersion `r sprintf("%.2f", ts$dispersion)`). Relative to the MMT, the cumulative RR was
`r rr(ts$rr, "cold_1")` at the 1st percentile and `r rr(ts$rr, "heat_99")` at the 99th.
The case-crossover design gave `r rr(cc$rr, "cold_1")` and `r rr(cc$rr, "heat_99")`.

```{r, out.width="85%"}
fig("fig1_overall_two_designs.png")
```

```{r}
knitr::kable(cc$compare, caption = "Table 2. Overall cumulative RR (95% CI) versus the time-series MMT, by design.")
```

### 4.2 Where and when the risk sits

Cold effects are delayed and persistent; heat effects are immediate and followed by a period of
RR below 1, the signature of short-term mortality displacement ("harvesting").

```{r, out.width="60%"}
fig("fig2_surface_3d.png")
```

```{r, out.width="85%"}
fig("fig3_surface_contour.png")
fig("fig4_lag_response.png")
```

### 4.3 Attributable burden

Non-optimal temperature accounted for `r afr(1)` of deaths: `r afr(2)` from cold and `r afr(3)` from
heat. Most of the burden came from *moderate* cold, not from extreme days.

```{r}
knitr::kable(af$af_print, caption = "Table 3. Attributable fraction (%) and deaths, with empirical 95% CIs.")
```

### 4.4 Sensitivity

```{r, out.width="95%"}
fig("fig5_sensitivity.png")
```

```{r}
knitr::kable(se$sens_print, caption = "Table 4. Sensitivity grid. Delta-QAIC is relative to the best-fitting model.")
```

Two honest observations. First, QAIC was lowest at `r best$df_per_year` df per year and lag
`r best$max_lag`, the edge of the df grid: QAIC keeps rewarding tighter seasonal control, which is
why the main specification was fixed beforehand. Second, the cold estimate is stable across the
grid (RR `r sprintf("%.2f to %.2f", min(se$sens$rr_cold), max(se$sens$rr_cold))`), whereas the heat
estimate is not (RR `r sprintf("%.2f to %.2f", min(se$sens$rr_heat), max(se$sens$rr_heat))`), because the
cumulative heat effect depends on how much of the post-heat deficit the lag window includes.

### 4.5 Heat x ozone interaction (warm season)

`r format(it$n_days, big.mark = ",")` warm-season days, `r it$n_high` with high ozone (lag 0-1 mean above
`r sprintf("%.0f", it$o3_cut)` ppb). Heat = `r sprintf("%.1f", it$t_heat)` C, reference = `r sprintf("%.1f", it$t_ref)` C.
The two heat surfaces differed (Wald chi-square `r sprintf("%.1f", it$wald)` on `r it$wald_df` df,
p `r format.pval(it$p_wald, digits = 2, eps = 0.001)`). The heat RR was `r im("RR10_heat_lowO3")` on low-ozone days and
`r im("RR_heat_within_highO3")` on high-ozone days. Multiplicative interaction: `r im("multiplicative_interaction")`;
RERI: `r im("RERI")`.

```{r, out.width="85%"}
fig("fig6_heat_by_ozone.png")
```

```{r}
knitr::kable(it$print, caption = "Table 5. Joint effects and interaction measures, with empirical 95% CIs.")
```

## 5. Limitations

- One city, 1987-2000, all-cause mortality: a teaching dataset, not a current population.
- Heat estimates lean heavily on the July 1995 heat wave (the series maximum, `r max(dat$death)` deaths
  on `r dat$date[which.max(dat$death)]`), which also coincided with high ozone.
- Exposure is a single city-wide daily mean (ecological); individual exposure is unmeasured.
- Dry-bulb temperature only. Humidity-inclusive indices (WBGT, UTCI) may describe heat strain better;
  that is what the HeatShield Agri forecasting layer computes, and comparing them is a natural next step.
- No control for influenza epidemics or PM10 (`r sum(is.na(dat$pm10))` missing days); relative humidity not modelled.
- The MMT is estimated with real uncertainty, and every RR and attributable fraction is conditional on it.
- In this dataset ozone and PM10 are approximate reconstructions of the original series (see
  `data/README.md`). The interaction analysis also dichotomises ozone and takes ozone status on the
  day of death while heat is cumulated over lags. It illustrates the method; it is not a causal estimate.

## 6. References

1. Gasparrini A, Armstrong B, Kenward MG. Distributed lag non-linear models. *Stat Med* 2010;29:2224-34. doi:10.1002/sim.3940
2. Gasparrini A. Distributed lag linear and non-linear models in R: the package dlnm. *J Stat Softw* 2011;43(8):1-20. doi:10.18637/jss.v043.i08
3. Armstrong BG, Gasparrini A, Tobias A. Conditional Poisson models: a flexible alternative to conditional logistic case cross-over analysis. *BMC Med Res Methodol* 2014;14:122. doi:10.1186/1471-2288-14-122
4. Gasparrini A, Leone M. Attributable risk from distributed lag models. *BMC Med Res Methodol* 2014;14:55. doi:10.1186/1471-2288-14-55
5. Gasparrini A, Guo Y, Hashizume M, et al. Mortality risk attributable to high and low ambient temperature: a multicountry observational study. *Lancet* 2015;386:369-75. doi:10.1016/S0140-6736(14)62114-0
6. Bhaskaran K, Gasparrini A, Hajat S, Smeeth L, Armstrong B. Time series regression studies in environmental epidemiology. *Int J Epidemiol* 2013;42:1187-95. doi:10.1093/ije/dyt092

## 7. Session

```{r}
si <- sessionInfo()
cat(si$R.version$version.string, "\n")
for (p in c("dlnm", "gnm", "MASS", "tsModel")) cat(sprintf("%-8s %s\n", p, as.character(packageVersion(p))))
```
